### 2023-02-13 talk/talk.tex
edit measles slides and start covid

Joint work with\\
Kidus Asfaw, Ning Ning, Joonha Park, Jesse Wheeler and Aaron King

@@

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?
  \item When should we want to?
  \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
  \end{enumerate}

@@

  \begin{enumerate}
\item Combining measurement noise and process noise.
\item Including covariates in mechanistically plausible ways.
\item  Continuous time models.
\item  Modeling and estimating interactions in coupled systems.
\item  Dealing with unobserved variables.
\item  \myemph{Modeling spatial-temporal dynamics}.
\item  Studying population dynamics via genetic sequence data.
  \end{enumerate}

  1--6 are from Bjornstad \& Grenfell ({\it Science}, 2001).\\
  7 is an active topic, but not the focus of this talk.\\
  1--5 are largely solved, from a methodological perspective.

@@

  \begin{itemize}
    \item We want to be able to fit arbitrary dynamic models. The limitations should be our scientific creativity and the information in the data.

    \item In practice, that means using \myemph{plug-and-play} methods which need a simulator from the model but not nice closed-form expressions for densities.

    \item We want statistically efficient inference, to extract all the information in the data.

    \item In practice, that means using likelihood-based methods.

@@

\begin{frame}{Parameters for the measles model}
  \bi
\item Seasonal tranmission: mean and amplitude, using school term for contact rate.
  \item Durations of latency and infectious period.
\item Initial values: fraction susceptible, latent and infectious.
\item Cohort effect: all births in an age cohort start school in September.
\item Inhomogenous mixing coefficient.
\item Measurement fraction.
\item Transport model gravity constant.
\item Dynamic noise (process overdispersion).
\item Measurement overdispersion.

@@

\begin{frame}{More on the block particle filter}

@@

\begin{columns}[T] % align columns
\begin{column}{.55\textwidth}
  \includegraphics[width=6cm]{slice_combined_plot-1.pdf}
\end{column}
\begin{column}{.52\textwidth}

@@

  Simulating $15$ year of data from $U=40$ cities for the measles model.
  Slice likelihood, varying $G$ with other paramters fixed at the truth.

@@

Results from Ionides et al (2021, {\it JASA}). We computed a slice due to lack of good optimization algorithms to compute a profile.

@@

@@

  \item Larger problems will need numerical approximations (e.g., EnKF). Exact Monte Carlo methods help study the effect of these approximations.

@@

\begin{columns}[T] % align columns
\begin{column}{.75\textwidth}
  \includegraphics[width=7.5cm]{ibpf/sim_search_boxplot-1.pdf}

@@

\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{5mm}

$\bullet$    IBPF applied to simulations (top panel) and data (bottom panel).

@@

$\bullet$    Multiple search refinements.

@@

    (A) Mostly shared parameters.

@@

    (B) All unit-specific parameters

@@

    (C) all unit-specific, without coupling

@@

@@

  \newcommand\futuresep{\vspace{3mm}}

  \begin{myitemize}
  \item We are reaching to the point where we can carry out likelihood-based inference for a flexible class of metapopulation models for measles.
Flexibility supports generation and testing of scientific hypotheses.

@@

%  \item Measles was previously a motivating model system for POMP methods for single populations.

@@

  \item Many systems in ecology, epidemiology and elsewhere could be studied in a SpatPOMP framework. Including microbiomes?

   \item Modeling and inference for nonlinear stochastic dynamics is hard. But, if you can't build a quantitative statistical model then you don't understand it and you can't control it?

>>>>

Joint work with\\
Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King

@@

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?

  \item When should we want to?

  \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
  \end{enumerate}

@@

  \begin{enumerate}
\item Combining measurement noise and process noise.
\item Including covariates in mechanistically plausible ways.
\item  Continuous time models.
\item  Modeling and estimating interactions in coupled systems.
\item  Dealing with unobserved variables.
\item  \myemph{Modeling spatiotemporal dynamics}.
\item  Studying population dynamics via genetic sequence data.
  \end{enumerate}

  1--5 are largely solved, from a methodological perspective.\\
  7 is exciting but not the focus of this talk.\\
  6 is our topic.

  \resizebox{12cm}{!}{Reviews: Bjornstad \& Grenfell ({\it Science}, 2001); Grenfell et al ({\it Science}, 2004)}

@@

  \begin{itemize}
    \item Consideration of arbitrary dynamic models. The limitations should be our scientific creativity and the information in the data.

     Hence, \myemph{plug-and-play} methods which need a simulator from the model but not nice closed-form expressions for densities.

    \item Statistically efficient inference, to extract all the information in the data.

    Hence, likelihood-based methods.

@@

\begin{frame}{Parameters for the measles model}
  \bi
\item Seasonal tranmission: mean and amplitude, using school term for contact rate.
  \item Durations of latency and infectious period.
\item Initial values: fraction susceptible, latent and infectious.
\item Cohort effect: all births in an age cohort start school in September.
\item Inhomogenous mixing coefficient.
\item Measurement fraction.
\item Population movement model: gravity constant.
\item Dynamic noise (process overdispersion).
\item Measurement overdispersion.

@@

\begin{frame}{More on the block particle filter (BPF)}

@@

\begin{columns}[T] % align columns
\begin{column}{.55\textwidth}
  \includegraphics[width=6cm]{slice_combined_plot-1.pdf}
\end{column}

\hspace{-5mm}
\begin{column}{.52\textwidth}

@@

  Simulating $15$ year of data from $U=40$ cities.
  Slice likelihood, varying $G$ with other parameters fixed.\\
   (Ionides et al, {\it JASA}, 2021).

@@

  \vspace{3mm}
  A much easier task than computing a profile likelihood for real data.

@@

\begin{frame}{Comments on the Ensemble Kalman Filter (EnKF)}

  \begin{itemize}
  \item EnKF is more dependent on approximate Gaussianity than is sometimes supposed.

  \item The Gaussian-inspired update rule is similar to the extended Kalman filter (EKF), which has largely been superseded by particle filter methods for low-dimensional nonlinear biological dynamics.

\item Simple systems can defeat EnKF: the linear Gaussian update is helpless when data inform the conditional variance rather than the conditional mean.

    \item Big systems need computationally tractable analysis. EnKF may sometimes be the best solution available, but be aware of its limitations.

@@

  \item Larger problems will need numerical approximations (e.g., EnKF).

@@

\begin{columns}[T] % align columns
\hspace{-4mm}\begin{column}{.75\textwidth}
  \includegraphics[width=7.5cm]{ibpf/sim_search_boxplot-1.pdf}

@@

\hspace{-18mm}
\begin{column}{.4\textwidth}
  \vspace{5mm}

  {\bf IBPF for measles, 20 cities}\\

  $\bullet$ simulations (top panel)\\
  dash line: likelihood at truth

  $\bullet$ data (bottom panel)\\
  dash line: MLE for case C.

@@

%$\bullet$    Multiple search refinements.

%    \vspace{5mm}

    (A) 9 shared parameters, $4\times 20$ unit-specific parameters

@@

    (B) all parameters unit-specific

@@

    (C) all unit-specific, with constant immigration instead of coupling

\begin{frame}{Inferences from the previous slide}

  \begin{itemize}
    %$\bullet$
  \item
IBPF can work for simulated data, since it finds likelihoods higher than the truth when the model is well-specified

@@

  \item
    %$\bullet$
    When coupled by a gravity model, simulated data show evidence against a constant immigration model.

  \item 
    %$\bullet$
    Real data do not favor this gravity model over a constant immigration model.
\end{itemize}

\begin{frame}{He et al (2010) parameters; $r$ is rank correlation with $N_{1950}$}

\resizebox{12cm}{!}{
\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r}
\hline
  & $N_{1950}$ & $R_0$ & IP & LP & $\alpha$ & $a$ & $\iota$ & $\psi$ & $\rho$ & $\sigma_{SE}$\\
\hline
Halesworth & 2200 & 33.00 & 2.30 & 7.90 & 0.95 & 0.38 & 0.0091 & 0.64 & 0.75 & 0.075\\
\hline
Lees & 4200 & 30.00 & 2.10 & 8.50 & 0.97 & 0.15 & 0.0310 & 0.68 & 0.61 & 0.080\\
\hline
Mold & 6400 & 21.00 & 1.80 & 5.90 & 1.00 & 0.27 & 0.0140 & 2.90 & 0.13 & 0.054\\
\hline
Dalton in Furness & 11000 & 28.00 & 2.00 & 5.50 & 0.99 & 0.20 & 0.0390 & 0.82 & 0.46 & 0.078\\
\hline
Oswestry & 11000 & 53.00 & 2.70 & 10.00 & 1.00 & 0.34 & 0.0300 & 0.48 & 0.63 & 0.070\\
\hline
Northwich & 18000 & 30.00 & 3.00 & 8.50 & 0.95 & 0.42 & 0.0600 & 0.40 & 0.80 & 0.086\\
\hline
Bedwellty & 29000 & 25.00 & 3.00 & 6.80 & 0.94 & 0.16 & 0.0400 & 0.95 & 0.31 & 0.061\\
\hline
Consett & 39000 & 36.00 & 2.70 & 9.10 & 1.00 & 0.20 & 0.0730 & 0.41 & 0.65 & 0.071\\
\hline
Hastings & 66000 & 34.00 & 5.40 & 7.00 & 1.00 & 0.30 & 0.1900 & 0.40 & 0.70 & 0.096\\
\hline
Cardiff & 240000 & 34.00 & 3.10 & 9.90 & 1.00 & 0.22 & 0.1400 & 0.27 & 0.60 & 0.054\\
\hline
Bradford & 290000 & 32.00 & 3.40 & 8.50 & 0.99 & 0.24 & 0.2400 & 0.19 & 0.60 & 0.045\\
\hline
Hull & 300000 & 39.00 & 5.50 & 9.20 & 0.97 & 0.22 & 0.1400 & 0.26 & 0.58 & 0.064\\
\hline
Nottingham & 310000 & 23.00 & 3.70 & 5.70 & 0.98 & 0.16 & 0.1700 & 0.26 & 0.61 & 0.038\\
\hline
Bristol & 440000 & 27.00 & 4.90 & 6.20 & 1.00 & 0.20 & 0.4400 & 0.20 & 0.63 & 0.039\\
\hline
Leeds & 510000 & 48.00 & 11.00 & 9.50 & 1.00 & 0.27 & 1.2000 & 0.17 & 0.67 & 0.078\\
\hline
Sheffield & 520000 & 33.00 & 6.40 & 7.20 & 1.00 & 0.31 & 0.8500 & 0.18 & 0.65 & 0.043\\
\hline
Manchester & 700000 & 33.00 & 6.90 & 11.00 & 0.96 & 0.29 & 0.5900 & 0.16 & 0.55 & 0.055\\
\hline
Liverpool & 800000 & 48.00 & 9.80 & 7.90 & 0.98 & 0.30 & 0.2600 & 0.14 & 0.49 & 0.053\\
\hline
Birmingham & 1100000 & 43.00 & 12.00 & 8.50 & 1.00 & 0.43 & 0.3400 & 0.18 & 0.54 & 0.061\\
\hline
London & 3400000 & 57.00 & 13.00 & 13.00 & 0.98 & 0.55 & 2.9000 & 0.12 & 0.49 & 0.088\\
\hline
$\qquad r$ & 1 & 0.46 & 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
\hline
\end{tabular}
}

@@

% From Li et al (2020) page S-14:
% According to official report, around 5 million people left Wuhan city before January 23. The total population in Wuhan city after January 23 (when travel restrictions were imposed) is therefore around 6 million.

\begin{frame}{COVID in 375 cities in China, 10-23 January, 2020}

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness, using an iterated EnKF \\
    (Li et al, {\it Science}, May 2020).
   \item
    The time interval predates Wuhan travel restrictions.
    %
%     load("data/incidence.rda")
%    li20dat <- incidence[1:14,-1]
%     tot <- sort(apply(li20dat,2,sum))
%     sum(tot)         
   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
   \item 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
\end{itemize}

@@

  \newcommand\futuresep{\vspace{5mm}}

  \begin{myitemize}

  \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.

@@

  \item The \texttt{spatPomp} package facilitates data analysis.

@@

    \item Questions and comments?
